---
title: "COVID"
output:
  pdf_document: default
  html_document: default
date: "2025-02-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(fig.width = 12, fig.height = 8)
library(tidyverse)
library(stringr)
library(forecast)
```

```{r get_jhu_data, include=FALSE}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names <- 
c("time_series_covid19_confirmed_US.csv",  "time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_US.csv",  "time_series_covid19_deaths_global.csv")

urls <- str_c(url_in, file_names)

global_cases <- read_csv(urls[2])
global_deaths <- read_csv(urls[4])
US_cases <- read_csv(urls[1])
US_deaths <- read_csv(urls[3])
```

After looking at global_cases and global_death, I would like to 
tidy those datasets and put each variable (date, cases, deaths)
in their own column.

Also I do not need Lat and Long for the analysis I am planning,
so I will get rid of those and rename Region and State to be more R friendly

```{r tidy_global_data, include=FALSE}
global_cases <- global_cases %>%
  pivot_longer(cols = -c('Province/State',
                         'Country/Region', Lat, Long),
                         names_to = "date",
                         values_to = "cases") %>%
  select(-c(Lat,Long))

global_deaths <- global_deaths %>%
  pivot_longer(cols = -c('Province/State',
                         'Country/Region', Lat, Long),
                          names_to = "date",
                          values_to = "deaths") %>%
  select(-c(Lat, Long))

global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = 'Country/Region',
         Province_State = 'Province/State') %>%
  mutate(date = mdy(date))

global <- global %>% filter(cases > 0)

global <- global %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)

uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

global <- global %>%
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths,
         Population, Combined_Key)
```

After looking at the global data we need to perform a similar process to the US data.

```{r tidy_US_data, include=FALSE}
US_cases <- US_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

US_deaths <- US_deaths %>%
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

US <- US_cases %>%
  full_join(US_deaths)
```

Now we are going to do some basic analysis of the data sets. First we will look at the deaths per million of each the country as a whole and then the deaths per million of Colorado by itself. 

```{r analyze_data, echo=FALSE, results='hide'}
US_by_state <- US %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Province_State, Country_Region, date, cases, deaths,
         deaths_per_mill, Population) %>%
  ungroup()

US_totals <- US_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  ungroup()

US_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) + 
  geom_point(aes(color = "cases")) + 
  geom_line(aes(y = deaths, color = "deaths")) + 
  geom_point(aes(y = deaths, color = "deaths")) + 
  scale_y_log10() + 
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in US", y = NULL)

  state <- "Colorado"
  US_by_state %>%
    filter(Province_State == state) %>%
    filter(cases > 0) %>%
    ggplot(aes(x = date, y = cases)) +
    geom_line(aes(color = "cases")) + 
    geom_point(aes(color = "cases")) + 
    geom_line(aes(y = deaths, color = "deaths")) + 
    geom_point(aes(y = deaths, color = "deaths")) + 
    scale_y_log10() + 
    theme(legend.position = "bottom",
          axis.text.x = element_text(angle = 90)) +
    labs(title = str_c("COVID19 in ", state), y = NULL)
  
```

Now we are going to look at the relationship between new cases and new deaths, both in the entire country and the State of Colorado.

```{r more_analysis, echo=FALSE, results='hide'}
  
  US_by_state <- US_by_state %>%
    mutate(new_cases = cases - lag(cases),
           new_deaths = deaths - lag(deaths))
  US_totals <- US_totals %>%
    mutate(new_cases = cases - lag(cases),
           new_deaths = deaths - lag(deaths))
  
  US_totals %>%
    ggplot(aes(x = date, y = new_cases)) + 
    geom_line(aes(color = "new_cases")) + 
    geom_point(aes(color = "new_cases")) + 
    geom_line(aes(y = new_deaths, color = "new_deaths")) + 
    geom_point(aes(y = new_deaths, color = "new_deaths")) + 
    scale_y_log10() + 
    theme(legend.position = "bottom",
          axis.text.x = element_text(angle = 90)) + 
    labs(title = "COVID19 in US", y = NULL)
  
  state <- "Colorado"
  US_by_state %>%
    filter(Province_State == state) %>%
    ggplot(aes(x = date, y = new_cases)) + 
    geom_line(aes(color = "new_cases")) + 
    geom_point(aes(color = "new_cases")) + 
    geom_line(aes(y = new_deaths, color = "new_deaths")) + 
    geom_point(aes(y = new_deaths, color = "new_deaths")) + 
    scale_y_log10() + 
    theme(legend.position = "bottom",
          axis.text.x = element_text(angle = 90)) + 
    labs(title = str_c("COVID19 in ", state),  y = NULL)
  
  US_state_totals <- US_by_state %>%
    group_by(Province_State) %>%
    summarize(deaths = max(deaths), cases = max(cases),
              population = max(Population),
              cases_per_thou = cases * 1000 / population,
              deaths_per_thou = deaths * 1000 / population) %>%
    filter(cases > 0, population > 0)
  
  US_state_totals %>%
    slice_min(deaths_per_thou, n = 10)
``` 

Now we are going to model some data. We are using a simple linear model so we will probaly not get the best fit.

```{r covid_modelling, echo=FALSE, results='hide'}
mod <- lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)

summary(mod)

US_state_totals %>%
  slice_min(cases_per_thou)

US_state_totals %>%
  slice_max(cases_per_thou)

US_state_totals %>% mutate(pred = predict(mod))

US_total_w_pred <- US_state_totals %>% mutate(pred = predict(mod))

US_total_w_pred %>% ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red")

```

I am interested in seeing what states had the highest prevalence of COVID cases in both total cases and cases per thousand.

```{r COVID_prevalence, echo=FALSE, results='hide', fig.height=12, fig.width=10}
avg_cases <- mean(US_state_totals$cases)

avg_data <- data.frame(Province_State = "Average", cases = avg_cases)

US_state_totals %>%
  ggplot(aes(x = reorder(Province_State, cases), y = cases, fill = Province_State)) +
  geom_col() +
  geom_text(aes(label = cases), position = position_dodge(width = 1), hjust= -0.1) +
  labs(title = "Total Cases by State",
       x = "Number of Cases",
       y = "State or Territory") +
  theme_minimal() +
  coord_flip() +
  geom_col(data = avg_data, aes(x = "Average", y = avg_cases), fill = "black") + 
  theme(legend.position = "none")

avg_cases_per_thou <- mean(US_state_totals$cases_per_thou)

avg_per_thou_data <- data.frame(Province_State = "Average", cases_per_thou = avg_cases_per_thou)
  
US_state_totals %>%
  ggplot(aes(x = reorder(Province_State, cases_per_thou), y = cases_per_thou, fill = Province_State)) +
  geom_col() +
  geom_text(aes(label = round(cases_per_thou, 0)), position = position_dodge(width = 0.5), hjust = -0.1) +
  labs(title = "Total Cases per Thousand by State",
       x = "Number of Cases per Thousand",
       y = "State or Territory") +
  theme_minimal() +
  coord_flip() +
  geom_col(data = avg_per_thou_data, aes(x = "Average", y = avg_cases_per_thou), fill = "black") +
  theme(legend.position = "none")
```

As you can see some states like Rhode Island and Alaska, as well as territories like Guam were disproportionately affected by COVID. 

I will now convert the data to time series in order to use the ARIMA model to predict cases for future years.

```{r arima_predictions, echo=FALSE}
US_totals_ts <- ts(US_totals$cases,
                   start = c(year(min(US_totals$date)), month(min(US_totals$date)), day(min(US_totals$date))),
                   frequency = 365)

fit_arima <- auto.arima(US_totals_ts)

forecast_rest_2023 <- forecast(fit_arima, h = 297)

autoplot(forecast_rest_2023) +
  labs(title = "US COVID Forecast for rest of 2023 (After 03/09/2023)",
       x = "Year",
       y = "Number of Cases") +
  theme_minimal()

```

This model is more than likely a better interpretation of the data than our previous linear model. The dark blue shaded region represents an 80% confidence interval while the lighter blue represents a 95% confidence interval.

```{r arima_deaths_predictions, echo=FALSE}
US_deaths_ts <- ts(US_totals$deaths,
                   start = c(year(min(US_totals$date)), month(min(US_totals$date)), day(min(US_totals$date))),
                   frequency = 365)

fit_arima_deaths <- auto.arima(US_deaths_ts)

forecast_deaths_2023 <- forecast(fit_arima_deaths, h = 297)

autoplot(forecast_deaths_2023) + 
  labs(title = "US COVID Deaths Forecast for rest of 2023 (After 03/09/2023",
       x = "Year",
       y = "Number of Deaths") +
  theme_minimal()
```